---
title: "Temporal persistence 2014:2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 10, cache = TRUE, cache.lazy = FALSE)
```

## Parameters
```{r}
lonLim   <- c(-9.5, 4.3)
latLim   <- c(36, 44)
datasets <- c("ERA5Land", "CIDE", "HARMONIE")
datasets_labels <- c("ERA5-Land", "CIDE", "HARMONIE-AROME")
years    <- list(ERA5 = 2014:2021, CIDE = 2014:2021, HARMONIE = 2014:2021)
```

## Library
```{r, cache=FALSE, message=FALSE, results='hide'}
library(loadeR)
library(visualizeR)
library(transformeR)
library(sp)
library(raster)
library(terra)
source("load_data.R")
```

## Region clusters
```{r}
clusters_sp <- as(vect("data/cluster_regions.shp"), "Spatial")
sp_layout   <- list(list(clusters_sp, first = FALSE, col = "black", lwd = 1.5))

make_multigrid <- function(clist, labels = datasets_labels) {
  if (!is.null(labels)) {
    for (i in seq_along(clist)) {
      if (!is.null(clist[[i]]$Variable)) {
        attr(clist[[i]]$Variable, "longname") <- labels[i]
      }
    }
  }
  do.call(makeMultiGrid, c(clist, list(skip.temporal.check = TRUE)))
}

grid_metrics_vs_ref <- function(grid, ref) {
  x <- as.vector(grid$Data)
  y <- as.vector(ref$Data)
  ok <- is.finite(x) & is.finite(y)

  if (!any(ok)) {
    return(c(rho = NA_real_, mae = NA_real_))
  }

  c(
    rho = suppressWarnings(cor(x[ok], y[ok], method = "spearman")),
    mae = mean(abs(x[ok] - y[ok]))
  )
}

label_with_metrics <- function(label, grid, ref, digits = 2) {
  metrics <- grid_metrics_vs_ref(grid, ref)
  sprintf("ρs = %.2f | MAE = %.2f",
          label, digits, metrics[["rho"]], digits, metrics[["mae"]])
}

label_with_metrics <- function(label, grid, ref, digits = 2) {
  metrics <- grid_metrics_vs_ref(grid, ref)
  sprintf("%s: \u03c1s = %.*f | MAE = %.*f",
          label, digits, metrics[["rho"]], digits, metrics[["mae"]])
}
```

## Integral timescale of the ACF
```{r tau_function}
tau.1D <- function(ts, lag.max = 30) {
  if (all(is.na(ts))) return(NA)
  ts <- ts[!is.na(ts)]
  if (length(ts) <= lag.max) return(NA)

  rho <- acf(ts, lag.max = lag.max, plot = FALSE)$acf[-1]

  idx0 <- which(rho <= 0)[1]
  if (!is.na(idx0) && idx0 > 1) {
    tau <- sum(rho[1:(idx0 - 1)])
  } else if (!is.na(idx0)) {
    tau <- rho[1]
  } else {
    tau <- sum(rho)
  }
  return(tau)
}
```

## Load data
```{r}
dat <- setNames(lapply(datasets, function(ds) {
  load_data(year = years[[ds]], lonLim = lonLim, latLim = latLim,
            what = "FWI", dataset = ds)
}), datasets)
```

## ACF integral timescale (tau)
```{r}
dat_tau <- lapply(dat, climatology, clim.fun = list(FUN = tau.1D))
```

```{r tau_multigrid, fig.width=14, fig.height=5, dpi=300, fig.path="figures/"}
tau_labels <- c(
  label_with_metrics("ERA5-Land", dat_tau$ERA5Land, dat_tau$CIDE),
  "CIDE (reference)",
  label_with_metrics("HARMONIE-AROME", dat_tau$HARMONIE, dat_tau$CIDE)
)
spatialPlot(make_multigrid(dat_tau, tau_labels), backdrop.theme = "countries", color.theme = "YlOrRd",
            at = seq(0, ceiling(max(sapply(dat_tau, \(x) max(x$Data, na.rm = TRUE)))), by = 0.5),
            main = expression(tau[30]), layout = c(3,1), sp.layout = sp_layout)
```

## Daily anomalies
```{r anomalies}
dat_anomalies <- lapply(dat, scaleGrid, time.frame = "daily")
```

## ACF integral timescale (tau) (anomalies)
```{r}
dat_tau <- lapply(dat_anomalies, climatology, clim.fun = list(FUN = tau.1D))
```

```{r tau_multigrid_anomalies, fig.width=16, fig.height=5, dpi=300, fig.path="figures/"}
tau_anom_labels <- c(
  label_with_metrics("ERA5-Land", dat_tau$ERA5Land, dat_tau$CIDE),
  "CIDE (reference)",
  label_with_metrics("HARMONIE-AROME", dat_tau$HARMONIE, dat_tau$CIDE)
)
spatialPlot(make_multigrid(dat_tau, tau_anom_labels), backdrop.theme = "countries", color.theme = "YlOrRd",
            at = seq(0, ceiling(max(sapply(dat_tau, \(x) max(x$Data, na.rm = TRUE)))), by = 0.5),
            main = expression(tau[30]), layout = c(3,1), sp.layout = sp_layout)
```

## Differences
```{r tau_diff_dat}
delta_era5_cide <- gridArithmetics(dat_tau$ERA5Land, dat_tau$CIDE, operator = "-")
delta_harmonie_cide <- gridArithmetics(dat_tau$HARMONIE, dat_tau$CIDE, operator = "-")

delta_dat_tau <- list(ERA5Land_CIDE = delta_era5_cide, HARMONIE_CIDE = delta_harmonie_cide)
```

```{r tau_diff, fig.width=14, fig.height=5, dpi=300, fig.path="figures/"}
max <- ceiling(max(sapply(delta_dat_tau, \(x) max(abs(x$Data), na.rm = TRUE))))
spatialPlot(make_multigrid(delta_dat_tau, c(
              label_with_metrics("ERA5-Land - CIDE", dat_tau$ERA5Land, dat_tau$CIDE),
              label_with_metrics("HARMONIE-AROME - CIDE", dat_tau$HARMONIE, dat_tau$CIDE)
            )), 
            backdrop.theme = "countries", color.theme = "RdBu", rev.colors = TRUE,
            at = seq(-max, max, 0.25), main = expression("Difference in " * tau[30]), layout = c(2,1), sp.layout = sp_layout)
```

# Spell

## Run-length statistics
```{r run_functions}
mean_run_yr <- function(x) {
  r    <- rle(x)
  runs <- r$lengths[r$values == 1]
  if (length(runs) == 0) return(0)
  mean(runs)
}

max_run_yr <- function(x) {
  r    <- rle(x)
  runs <- r$lengths[r$values == 1]
  if (length(runs) == 0) return(0)
  max(runs)
}
```

## P95 excedance
```{r excedance}
p95 <- lapply(dat, climatology,
              clim.fun = list(FUN = "quantile", probs = 0.95, na.rm = TRUE))

dat_exc <- setNames(lapply(datasets, function(ds) {
  out          <- dat[[ds]]
  out$Data     <- 1L * sweep(dat[[ds]]$Data, c(2, 3), p95[[ds]]$Data, FUN = ">")
  attributes(out$Data) <- attributes(dat[[ds]]$Data)
  out
}), datasets)
```

## Annual run-length aggregation
```{r run_aggregate}
dat_mean_run <- lapply(dat_exc, aggregateGrid, aggr.y = list(FUN = mean_run_yr))
dat_max_run  <- lapply(dat_exc, aggregateGrid, aggr.y = list(FUN = max_run_yr))
```

## Climatological summaries
```{r run_climatology}
clim_mean <- lapply(dat_mean_run, climatology, clim.fun = list(FUN = mean, na.rm = TRUE))
clim_sd   <- lapply(dat_mean_run, climatology, clim.fun = list(FUN = sd,   na.rm = TRUE))
clim_max  <- lapply(dat_max_run,  climatology, clim.fun = list(FUN = mean, na.rm = TRUE))
```

## Mean annual event duration
```{r mean_run_multigrid, fig.width=16, fig.height=5, dpi=300, fig.path="figures/", warning=FALSE}
mean_run_grid <- list(ERA5Land = clim_mean$ERA5Land, CIDE = clim_mean$CIDE, HARMONIE = clim_mean$HARMONIE)
mean_run_labels <- c(
  label_with_metrics("ERA5-Land", clim_mean$ERA5Land, clim_mean$CIDE),
  "CIDE (reference)",
  label_with_metrics("HARMONIE-AROME", clim_mean$HARMONIE, clim_mean$CIDE)
)
spatialPlot(make_multigrid(mean_run_grid, mean_run_labels), backdrop.theme = "countries", color.theme = "YlOrRd",
            at = seq(0, ceiling(max(sapply(mean_run_grid, \(x) max(x$Data, na.rm = TRUE)))), by = 0.5),
            main = "Climatology of mean annual FWI>P95 event duration (days)", layout = c(3,1), sp.layout = sp_layout)
```

## SD of spell length
```{r sd_run_multigrid, fig.width=16, fig.height=5, dpi=300, fig.path="figures/", warning=FALSE}
sd_run_grid <- list(ERA5Land = clim_sd$ERA5Land, CIDE = clim_sd$CIDE, HARMONIE = clim_sd$HARMONIE)
sd_run_labels <- c(
  label_with_metrics("ERA5-Land", clim_sd$ERA5Land, clim_sd$CIDE),
  "CIDE (reference)",
  label_with_metrics("HARMONIE-AROME", clim_sd$HARMONIE, clim_sd$CIDE)
)
spatialPlot(make_multigrid(sd_run_grid, sd_run_labels), backdrop.theme = "countries", color.theme = "YlOrRd",
            at = seq(0, ceiling(max(sapply(sd_run_grid, \(x) max(x$Data, na.rm = TRUE)))), by = 0.5),
            main = "SD of annual mean FWI>P95 event duration (days)", layout = c(3,1), sp.layout = sp_layout)
```

## Max annual event duration
```{r max_run_multigrid, fig.width=16, fig.height=5, dpi=300, fig.path="figures/", warning=FALSE}
max_run_grid <- list(ERA5Land = clim_max$ERA5Land, CIDE = clim_max$CIDE, HARMONIE = clim_max$HARMONIE)
max_run_labels <- c(
  label_with_metrics("ERA5-Land", clim_max$ERA5Land, clim_max$CIDE),
  "CIDE (reference)",
  label_with_metrics("HARMONIE-AROME", clim_max$HARMONIE, clim_max$CIDE)
)
spatialPlot(make_multigrid(max_run_grid, max_run_labels), backdrop.theme = "countries", color.theme = "YlOrRd",
            at = seq(0, ceiling(max(sapply(max_run_grid, \(x) max(x$Data, na.rm = TRUE)))), by = 0.5),
            main = "Climatology of maximum annual FWI>P95 event duration (days)", layout = c(3,1), sp.layout = sp_layout)
```

## Differences (ERA5 − CIDE)
```{r run_diff}
delta_mean <- list(ERA5Land = gridArithmetics(clim_mean$ERA5Land, clim_mean$CIDE, operator = "-"),
                   HARMONIE = gridArithmetics(clim_mean$HARMONIE, clim_mean$CIDE, operator = "-"))
delta_sd   <- list(ERA5Land = gridArithmetics(clim_sd$ERA5Land,   clim_sd$CIDE,   operator = "-"),
                   HARMONIE = gridArithmetics(clim_sd$HARMONIE,   clim_sd$CIDE,   operator = "-"))
delta_max  <- list(ERA5Land = gridArithmetics(clim_max$ERA5Land,  clim_max$CIDE,  operator = "-"),
                   HARMONIE = gridArithmetics(clim_max$HARMONIE,  clim_max$CIDE,  operator = "-"))
```

```{r diff_mean_run, fig.width=14, fig.height=5, dpi=300, fig.path="figures/", warning=FALSE}
spatialPlot(make_multigrid(delta_mean, c(
              label_with_metrics("ERA5-Land - CIDE", clim_mean$ERA5Land, clim_mean$CIDE),
              label_with_metrics("HARMONIE-AROME - CIDE", clim_mean$HARMONIE, clim_mean$CIDE)
            )), backdrop.theme = "countries",
            color.theme = "RdBu", rev.colors = TRUE,
            main      = "Difference in Climatology of mean annual FWI>P95 spell length (days)",
            sp.layout = sp_layout, layout = c(2,1))
```

```{r diff_sd_run, fig.width=14, fig.height=5, dpi=300, fig.path="figures/", warning=FALSE}
spatialPlot(make_multigrid(delta_sd, c(
              label_with_metrics("ERA5-Land - CIDE", clim_sd$ERA5Land, clim_sd$CIDE),
              label_with_metrics("HARMONIE-AROME - CIDE", clim_sd$HARMONIE, clim_sd$CIDE)
            )), backdrop.theme = "countries", color.theme = "RdBu", rev.colors = TRUE,
            main      = "Difference in Climatology of SD of annual mean FWI>P95 spell length (days)",
            sp.layout = sp_layout, layout = c(2,1))
```

```{r diff_max_run, fig.width=14, fig.height=5, dpi=300, fig.path="figures/", warning=FALSE}
spatialPlot(make_multigrid(delta_max, c(
              label_with_metrics("ERA5-Land - CIDE", clim_max$ERA5Land, clim_max$CIDE),
              label_with_metrics("HARMONIE-AROME - CIDE", clim_max$HARMONIE, clim_max$CIDE)
            )), backdrop.theme = "countries", color.theme = "RdBu", rev.colors = TRUE,
            main      = "Difference in Climatology of maximum annual FWI>P95 event duration (days)",
            sp.layout = sp_layout, layout = c(2,1))
```

